---
title: "*A five-decision testing procedure*"
---

<style type="text/css">
body{ font-size: 16px; }
</style>

&nbsp;

<div style="border:1px solid black ; text-align:center">

# [Aaron McDaid](https://aaronmcdaid.github.io/)

Statistics and C++, and other fun projects

`aaron.mcdaid@gmail.com`

[`@aaronmcdaid`](https://twitter.com/aaronmcdaid)
</div>

# &nbsp;

<div style="text-align:right">
***by [Aaron](https://aaronmcdaid.github.io/), 30th October 2017***

[(code for this document)](index.Rmd)
</div>

# &nbsp;

*This is a short post summarising some recent work
[(**arXiv link**)](https://arxiv.org/abs/1710.06676)
I did with [Zoltan Kutalik](https://twitter.com/zkutalik)
and [Valentin Rousson](https://www.iumsp.ch/fr/rousson-valentin)*

# The null hypothesis

Given a sample of, say, $n=100$ observations from a Normal distribution with variance 100,
we wish to make inferences about the mean, $\theta$,

$$ x_i \sim \mathcal{N}(\theta,100) $$

For simplicity, assume that we really just want to know which is these three are true:

1. $\theta < 0$
1. $\theta = 0$
1. $\theta > 0$

We compute the sample mean from our 100 observations,

$$ \bar{x} = \frac1n \sum_{i=1}^n x_i $$

If the sample mean is large, e.g. $\bar{x} > 10$, then it is virtually certain that $\theta > 0$ and we
can safely *reject* the other two hypotheses.
This is because, if $\theta{\le}0$, the probability of seeing $\bar{x} > 10$ is very small,
less than $7 \times 10^{-24}$.

```{r echo=F}
# PP for Pretty Print. Print a variables name and value, separated by '='
# You can pass in multiple values simultaneously
PP <- function(x,...) {
	cat(format(width=30, justify="right", deparse(substitute(x))))
	cat('  =  ')
	catn(deparse(x));
	if(length(list(...))>0) {
		PP(...)
	}
}
```

```{r}
variance.of.one.observation = 100
n=100
variance.of.sample.mean = variance.of.one.observation / n
prob.of.sample.mean.at.least.10 = pnorm(10, mean=0, sd=sqrt(variance.of.sample.mean), lower.tail=F)


PP(n, variance.of.sample.mean, prob.of.sample.mean.at.least.10)
```

Similarly, if $\bar{x} < -10$, then it is virtually certain that $\theta > 0$ and we
can safely *reject* the other two hypotheses.

# Error rates

If our sample mean is close to zero however, what to we do?
We need to set thresholds, where we accept our reject the hypotheses
depending on the value of the sample mean.
It is common to decide these thresholds by first fixing an error rate,
typically $\alpha = 0.05$, and then calculating the thresholds such that the probability
of incorrectly rejecting a hypothesis is no more than 5%.

We propose dividing the possible sample means into five bins.
As far as we are aware, this particular procedure has not been proposed before.
Depending on which bin the sample means falls into, which accept or reject
4according to this scheme, depending on where $\bar{x}$ is relative to the four
'special' values, $-1.96, -1.64, 1.64, 1.96$.
Look up your sample mean on the x-axis of this plot, and then look up
the reject/accept rule written in that window.
There are five windows in this plot, hence the "five-decision rule".

```{r}
outer.threshold = qnorm(0.025, mean=0, sd=sqrt(variance.of.sample.mean), lower.tail=F)
inner.threshold = qnorm(0.05 , mean=0, sd=sqrt(variance.of.sample.mean), lower.tail=F)
PP(outer.threshold, inner.threshold)
```

```{r echo=F}
plot(dnorm ,xlim=c(-3,3)
     ,xaxt='n'
     ,xlab='' ,ylab='', col='lightblue'
     )
axis(1, format(digits=4,c(-inner.threshold,-outer.threshold,inner.threshold,outer.threshold))
     ,las=2
     )
abline(v= outer.threshold, col='grey')
abline(v= inner.threshold, col='grey')
abline(v=-outer.threshold, col='grey')
abline(v=-inner.threshold, col='grey')
text (-2.5, 0.2
     , srt=-90
     ,labels = expression('Reject   '* {theta>=0} * '    Accept  '* {theta< 0} )
     )
text (-1.8, 0.2
     , srt=-90
     ,labels = expression('Reject   '* {theta> 0} * '    Accept  '* {theta<=0} )
     )
text (0   , 0.2
     , srt=-90
     ,labels = c('Reject   nothing,\naccept  everything')
     )
text (1.8, 0.2
     , srt=-90
     ,labels = expression('Reject   '* {theta< 0} * '    Accept  '* {theta>=0} )
     )
text (2.5, 0.2
     , srt=-90
     ,labels = expression('Reject   '* {theta<=0} * '    Accept  '* {theta> 0} )
     )
```

To prove that the error rate is at most 5%, we consider each of the three
possible "truths" in turn:

**Possibility 1/3: **
If $\theta<0$, then only the last two bins, when $\bar{x}>1.64$,
constitute an error. For any $\theta<0$,
the probability
of observing $\bar{x}>1.64$ is at most 5%.
Of course, for $\theta$ which is truly much smaller than zero, the
probability of error will be much smaller than 5%.

**Possibility 2/3: **
If $\theta=0$, then only the outer two bins, when $|\bar{x}|>1.96$,
constitute an error. And the probability, assuming $\theta=0$,
of observing $|\bar{x}|>1.96$ is 5%.

**Possibility 3/3: **
If $\theta>0$, then only the first two bins, when $\bar{x}<-1.64$,
constitute an error. For any $\theta>0$,
the probability
of observing $\bar{x}<-1.64$ is at most 5%.
Of course, for $\theta$ which is truly much greater than zero, the
probability of error will be much smaller than 5%.

&nbsp;

&nbsp;
<div style="text-align:right">
***by [Aaron](https://aaronmcdaid.github.io/), 29th October 2017***
</div>

<div id="disqus_thread"></div>
<script type="text/javascript">
    /* * * CONFIGURATION VARIABLES: EDIT BEFORE PASTING INTO YOUR WEBPAGE * * */
    var disqus_shortname = 'aaron-mcdaid'; // required: replace example with your forum shortname
    /* * * DON'T EDIT BELOW THIS LINE * * */
    (function() {
        var dsq = document.createElement('script'); dsq.type = 'text/javascript'; dsq.async = true;
        dsq.src = '//' + disqus_shortname + '.disqus.com/embed.js';
        (document.getElementsByTagName('head')[0] || document.getElementsByTagName('body')[0]).appendChild(dsq);
    })();
</script>
<noscript>Please enable JavaScript to view the <a href="http://disqus.com/?ref_noscript">comments powered by Disqus.</a></noscript>
